---
title: "Convert study template data and generate summary stats / plots"
author: "Biao Li"
date: "`r Sys.Date()`"
header-includes: 
 \usepackage{geometry}
 \geometry{top=0.75in,left=0.6in,bottom=0.75in,right=0.6in}
 \usepackage{titlesec}
 \titleformat{\chapter}{\normalfont\LARGE\bfseries}{\thechapter}{1em}{}
 \titlespacing*{\chapter}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
output:
  pdf_document: 
    toc: true
    number_sections: true
  html_document:
    code_folding: hide
    toc: true
  word_document: default
documentclass: report
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(echo=FALSE)
```


```{r pkg_}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(patchwork)
library(readxl)
library(plotrix)

## read TV_measurements tab from ../data/
dat_tv <- as.data.frame(read_excel("../data/in_vivo/MM_study5_data_example_revised_template.xlsm", sheet = "TV_measurements")) %>%
  select(-Width_mm, -Length_mm) %>%
  drop_na(Mouse_ID) %>%
  select(Mouse_ID, Group, TV, Day)


groups <- unique(dat_tv$Group)
days <- sort(unique(dat_tv$Day))


file_name <- "output.xlsx"
sheet_name <- " Prism TV"

# create workbook and add worksheet
wb <- createWorkbook()
addWorksheet(wb, sheet_name)

list_df_grp <- list()
n_rows <- c()
n_cols <- c()

row_to_start <- 2

# for each group create its df
for (idx in 1:length(groups)) {
  
  grp <- groups[idx]
  df_grp <- dat_tv %>%
    filter(Group == grp) %>%
    select(-Group) %>%
    pivot_wider(
      names_from = Mouse_ID,
      values_from = TV
    )
  
  list_df_grp[[grp]] <- df_grp
  n_rows <- c(n_rows, nrow(df_grp))
  n_cols <- c(n_cols, ncol(df_grp))
  
  
  # add datatable of each group
  writeDataTable(wb, sheet = sheet_name, 
                 x = df_grp, 
                 startRow = row_to_start, 
                 startCol = 1)
  
  
  # add merged cell of group name
  mergeCells(wb, sheet = sheet_name, cols = 1:n_cols[idx], rows = row_to_start - 1)
  writeData(wb, sheet = sheet_name, x = grp, startCol = 1, startRow = row_to_start - 1)
  
  # update row_to_start for table of next group
  row_to_start <- row_to_start + n_rows[idx] + 3 # +3 --> colname+1, space+1, groupname+1

}


# for all groups create one long table
# add a header of "All Groups"
writeData(wb, sheet = sheet_name, x = "All Groups", startCol = 1, startRow = row_to_start - 1)

for (idx in 1:length(groups)) {
  grp <- groups[idx]
  
  if (idx == 1) {
    
    col_to_start <- 2
    
    df_all <- list_df_grp[[grp]]
    
  } else {
    
    col_to_start <- col_to_start + n_cols[idx] - 1
    
    df_all <- df_all %>%
      bind_cols(list_df_grp[[grp]][, 2:n_cols[idx]])
  }
  
  # add merged cell of each group name
  mergeCells(wb, sheet = sheet_name, cols = col_to_start:(col_to_start + n_cols[idx] - 2), rows = row_to_start - 1)
  writeData(wb, sheet = sheet_name, x = grp, startCol = col_to_start, startRow = row_to_start - 1)
  
}

# add datatable of all groups
writeDataTable(wb, 
               sheet = sheet_name, 
               x = df_all, 
               startRow = row_to_start, 
               startCol = 1)


## add TV summary tab to workbook
sheet_name_tv_summary <- "TV Summary"
addWorksheet(wb, sheet_name_tv_summary)

# create header row 
writeData(wb, sheet = sheet_name_tv_summary, x = t(as.data.frame(c("Group", "Mouse_ID", paste0("Day_", days)))), 
          startCol = 1, startRow = 1, 
          colNames = FALSE)

# for each group output TV then followed by summary stat
row_to_start <- 2

for (idx in 1:length(groups)) {
  grp <- groups[idx]
  
  df_grp <- dat_tv %>%
    filter(Group == grp) %>%
    select(-Group) %>%
    pivot_wider(
      names_from = Day,
      values_from = TV
  )
  
  # write TV table df_grp
  writeDataTable(wb, sheet = sheet_name_tv_summary, 
                 x = df_grp, 
                 startRow = row_to_start, 
                 startCol = 2)
  
  # write merged cell of Group info
  mergeCells(wb, sheet = sheet_name_tv_summary,
             cols = 1,
             rows = row_to_start : (row_to_start + nrow(df_grp)))
  writeData(wb, sheet = sheet_name_tv_summary, x = grp, startCol = 1, startRow = row_to_start)

  row_to_start <- row_to_start + nrow(df_grp) + 6
  
  #   # add merged cell of each group name
  # mergeCells(wb, sheet = sheet_name, cols = col_to_start:(col_to_start + n_cols[idx] - 2), rows = row_to_start - 1)
  # writeData(wb, sheet = sheet_name, x = grp, startCol = col_to_start, startRow = row_to_start - 1)
  
  #   # add datatable of each group
  # writeDataTable(wb, sheet = sheet_name, 
  #                x = df_grp, 
  #                startRow = row_to_start, 
  #                startCol = 1)
  # 
  # 
  # 
  
}

## add n, mean, median, stddev and stderr to each group
dat_stat <- dat_tv %>%
  group_by(Group, Mouse_ID) %>%
  summarise(
    Mean = mean(TV, na.rm = TRUE),
    Median = median(TV, na.rm = TRUE),
    StdDev = sd(TV, na.rm = TRUE),
    StdErr = std.error(TV, na.rm = TRUE), 
    N = n()
  ) 

## save workbook
saveWorkbook(wb, file = file_name, overwrite = TRUE)


#####################
# 
# # add mean and sd to each group
# dat_stat <- dat %>%
#   group_by(Dox.condition, genotype, Day) %>%
#   summarise(
#     mean = mean(TV, na.rm = TRUE), 
#     median = median(TV, na.rm = TRUE), 
#     sd = sd(TV, na.rm = TRUE), 
#     mad = mad(TV, na.rm = TRUE), 
#     n = n()
#   ) %>%
#   mutate(factor = case_when(
#     Dox.condition == "Dox" ~ paste0(genotype, " + dox"), 
#     TRUE ~ paste0(genotype, " - dox")
#   ))



# dat <- read.csv("03-04-2024_CypAKD Model 5_KPCY_96 animals_COPY.csv", header = T) %>%
#   mutate(Genotype = Injected_KPCY_cells_Genotypes) %>%
#   mutate(Condition_Genotype_Operator = paste(Condition, Genotype, Operator, sep = "_")) %>%
#   mutate(Condition_Operator = paste(Condition, Operator, sep = "_")) %>%
#   mutate(Genotype_Operator = paste(Genotype, Operator, sep = "_")) %>%
#   mutate(Genotype_Condition = paste(Genotype, Condition, sep = "_"))
# 
# dat_new <- read.csv("03-04 and 06-2024_CypAKD Model 5_KPCY_96 animals_flat.csv", header = T) %>%
#   mutate(Genotype = Injected_KPCY_cells_Genotypes) %>%
#   mutate(Operator = Pseudo_Operator) %>%
#   mutate(Genotype_Operator = paste(Genotype, Pseudo_Operator, sep = "_")) %>%
#   mutate(Genotype_Condition = paste(Genotype, Condition, sep = "_")) %>%
#   mutate(Genotype_PseudoOperator = paste(Genotype, Pseudo_Operator, sep = "_"))
```



```{r test_revised_template, include=FALSE}
library(lubridate)
## add pseudo Date_treatment as 4/1/24 and pseudo (measurement)
dat_tmp <- dat %>%
  mutate(Date_treatment = format(mdy("04/01/2024"), "%m/%d/%y")) %>%
  mutate(Date = format(mdy(Date_treatment) + days(Day), "%m/%d/%y")) %>%
  mutate(Group = case_when(
    Dox.condition == "Dox" ~ paste0(genotype, " + Dox"), 
    .default = paste0(genotype, " - Dox")
  )) %>%
  mutate(Mouse_ID = Animal.ID) %>%
  arrange(Date) %>%
  select(Mouse_ID, Date, TV, Group, Day) 


write.csv(file = "study5_tv_for_testing_revised_data_template.csv", dat_tmp)

## list of Mouse IDs and their groups
dat_tmp_ids <- dat_tmp %>%
  distinct(Mouse_ID, .keep_all = TRUE) %>%
  arrange(Group, Mouse_ID)

write.csv(file = "study5_mouse_metadata_testing_revised_data_template.csv", dat_tmp_ids, row.names = F)
  

```


# Plot - all original data
```{r plot, fig.height = 8, fig.width = 6}
pal <- c("darkgray", "darkgray", "#00B159", "#00B159", "#00AEDB", "#00AEDB")
# ltype <- c("solid", "dashed", "solid", "dashed", "solid", "dashed")
ltype <- rep(c("dashed", "solid"), 3)

# ggplot(dat_stat, aes(Day, mean, group = factor, color = factor, linetype = factor)) +
#   geom_line() + 
#   geom_point(size = 2) + 
#   geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), 
#                 width = 4, 
#                 position = position_dodge(0.5)) + 
#   theme_classic() + 
#   xlim(0, NA) + 
#   scale_linetype_manual(values = ltype, name = "") + 
#   scale_color_manual(values = pal, name = "") + 
#   theme(legend.position = c(0.15, 0.8),
#         legend.key.width = unit(0.5, "inch"), 
#         axis.title.y = element_blank())


width <- 0.5

p <- ggplot(dat_stat, aes(Day, mean, group = factor, color = factor, linetype = factor)) +
  geom_line() +
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean - sd, yend = mean - sd 
  ), 
  size = 0.5, linetype = "solid") + 
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean + sd, yend = mean + sd
  ), 
  size = 0.5, linetype = "solid") + 
  geom_linerange(aes(
    x = Day, 
    ymin = (mean - sd), 
    ymax = (mean + sd)
  ), 
  size = 0.5) + 
  
  geom_point(size = 2) + 
  # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), 
  #               width = 4, 
  #               position = position_dodge(0.5)) + 
  theme_classic() + 
  xlim(0, NA) + 
  scale_linetype_manual(values = ltype, name = "") + 
  scale_color_manual(values = pal, name = "") + 
  theme(legend.position = c(0.35, 0.8),
        legend.key.height = unit(0.5, "inch"), 
        legend.key.width = unit(0.2, "inch"), 
        axis.title.y = element_blank()) + 
  guides(linetype = guide_legend(ncol = 3))

p_facet <- p + 
  facet_wrap(~ genotype) + 
  theme(
    legend.position = "none"
  )


p / p_facet



```

## summary stat - all original data {.tabset} 
```{r summary_stat, results = "asis"}

# table view of dat_stat

# dat_stat_tbl <- dat_stat %>%
#   mutate(txt = paste0(round(mean, 2), " (", round(sd, 2), ")\n", "n = ", n), .groups = "drop") %>%
#   select(factor, Day, txt) %>%
#   group_by(factor, Day) %>%
#   pivot_wider(names_from = factor, values_from = txt)
dat_stat_tbl <- dat_stat %>%
  as_tibble() %>%
  mutate(txt = paste0(round(mean, 2), " (", round(sd, 2), ") ", "n = ", n), .groups = "drop") %>%
  select(factor, Day, txt) %>%
  mutate(Day = paste0("Day ", Day)) %>%
  pivot_wider(names_from = Day, values_from = txt)

print(knitr::kable(dat_stat_tbl, format = "markdown"))

```


# Data profiling - proportion of TV > 0 {.tabset}

```{r profiling, results = "asis"}
days <- sort(unique(dat$Day))

for (d in days) {
  cat('\n##', paste0("Day ", d), "\n")
  dat_d <- dat %>%
    filter(Day == d) %>%
    group_by(Dox.condition, genotype) %>%
    summarize(percent_non_0 = round(mean(TV > 0, na.rm = TRUE), 3), .groups = 'drop') %>%
    pivot_wider(names_from = genotype, values_from = percent_non_0)
  print(knitr::kable(dat_d))
  cat("\n")
  
}

```

# Plot - only mice of TV > 0 at day 19 (for each mouse normalizing TV to its day 19 measurement)

```{r plot_mice_d19, fig.height = 8, fig.width = 6}
selected_mice <- dat %>%
  filter(Day == 19) %>%
  filter(TV > 0) %>%
  pull(Animal.ID) %>%
  unique()

# select only mice of non-0 TV at day 19 and only day 19 and onward
dat_s <- dat %>%
  filter(Day >= 19) %>%
  filter(Animal.ID %in% selected_mice)

# normalize TV to day 19 per each mouse
dat_sn <- dat_s %>%
  group_by(Animal.ID) %>%
  mutate_at(5, funs(./first(.)))

# stat of dat_sn
dat_sn_stat <- dat_sn %>%
  group_by(Dox.condition, genotype, Day) %>%
  summarise(
    mean = mean(TV, na.rm = TRUE), 
    median = median(TV, na.rm = TRUE), 
    sd = sd(TV, na.rm = TRUE), 
    mad = mad(TV, na.rm = TRUE), 
    n = n()
  ) %>%
  mutate(factor = case_when(
    Dox.condition == "Dox" ~ paste0(genotype, " + dox"), 
    TRUE ~ paste0(genotype, " - dox")
  ))


width <- 0.5

p_sn <- ggplot(dat_sn_stat, aes(Day, mean, group = factor, color = factor, linetype = factor)) +
  geom_line() +
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean - sd, yend = mean - sd 
  ), 
  size = 0.5, linetype = "solid") + 
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean + sd, yend = mean + sd
  ), 
  size = 0.5, linetype = "solid") + 
  geom_linerange(aes(
    x = Day, 
    ymin = (mean - sd), 
    ymax = (mean + sd)
  ), 
  size = 0.5) + 
  
  geom_point(size = 2) + 
  # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), 
  #               width = 4, 
  #               position = position_dodge(0.5)) + 
  theme_classic() + 
  xlim(0, NA) + 
  scale_linetype_manual(values = ltype, name = "") + 
  scale_color_manual(values = pal, name = "") + 
  theme(legend.position = c(0.35, 0.8),
        legend.key.height = unit(0.5, "inch"), 
        legend.key.width = unit(0.2, "inch"), 
        axis.title.y = element_blank()) + 
  guides(linetype = guide_legend(ncol = 3))

p_sn_facet <- p_sn + 
  facet_wrap(~ genotype) + 
  theme(
    legend.position = "none"
  ) + 
  xlim(min(dat_sn_stat$Day), NA)

p_sn / p_sn_facet


```


## summary stat - normalizing to day 19 {.tabset} 

```{r summary_stat_sn, results = "asis"}

dat_sn_stat_tbl <- dat_sn_stat %>%
  as_tibble() %>%
  mutate(txt = paste0(round(mean, 2), " (", round(sd, 2), ") ", "n = ", n), .groups = "drop") %>%
  select(factor, Day, txt) %>%
  mutate(Day = paste0("Day ", Day)) %>%
  pivot_wider(names_from = Day, values_from = txt)
  
print(knitr::kable(dat_sn_stat_tbl, format = "markdown"))

```

# Plot - all mice but removing TV == 0 measurements (for each mouse normalizing TV to its first non-zero measurement)

```{r plot_mice_nonzero, fig.height = 8, fig.width = 6}
# selected_mice <- dat %>%
#   filter(Day == 19) %>%
#   filter(TV > 0) %>%
#   pull(Animal.ID) %>%
#   unique()

# select only mice of non-0 TV 
dat_s <- dat %>%
  # filter(Day >= 19) %>%
  filter(TV > 0)

# normalize TV to day 19 per each mouse
dat_sn <- dat_s %>%
  group_by(Animal.ID) %>%
  mutate_at(5, funs(./first(.)))

# stat of dat_sn
dat_sn_stat <- dat_sn %>%
  group_by(Dox.condition, genotype, Day) %>%
  summarise(
    mean = mean(TV, na.rm = TRUE), 
    median = median(TV, na.rm = TRUE), 
    sd = sd(TV, na.rm = TRUE), 
    mad = mad(TV, na.rm = TRUE), 
    n = n()
  ) %>%
  mutate(factor = case_when(
    Dox.condition == "Dox" ~ paste0(genotype, " + dox"), 
    TRUE ~ paste0(genotype, " - dox")
  ))


width <- 0.5

p_sn <- ggplot(dat_sn_stat, aes(Day, mean, group = factor, color = factor, linetype = factor)) +
  geom_line() +
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean - sd, yend = mean - sd 
  ), 
  size = 0.5, linetype = "solid") + 
  geom_segment(aes(
    x = Day - width, xend = Day + width, 
    y = mean + sd, yend = mean + sd
  ), 
  size = 0.5, linetype = "solid") + 
  geom_linerange(aes(
    x = Day, 
    ymin = (mean - sd), 
    ymax = (mean + sd)
  ), 
  size = 0.5) + 
  
  geom_point(size = 2) + 
  # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), 
  #               width = 4, 
  #               position = position_dodge(0.5)) + 
  theme_classic() + 
  xlim(0, NA) + 
  scale_linetype_manual(values = ltype, name = "") + 
  scale_color_manual(values = pal, name = "") + 
  theme(legend.position = c(0.35, 0.8),
        legend.key.height = unit(0.5, "inch"), 
        legend.key.width = unit(0.2, "inch"), 
        axis.title.y = element_blank()) + 
  guides(linetype = guide_legend(ncol = 3))

p_sn_facet <- p_sn + 
  facet_wrap(~ genotype) + 
  theme(
    legend.position = "none"
  ) + 
  xlim(min(dat_sn_stat$Day), NA)

p_sn / p_sn_facet


```

## summary stat - normalizing TV of each mouse to its first non-zero measurement {.tabset} 

```{r summary_stat_nonzero, results = "asis"}

dat_sn_stat_tbl <- dat_sn_stat %>%
  as_tibble() %>%
  mutate(txt = paste0(round(mean, 2), " (", round(sd, 2), ") ", "n = ", n), .groups = "drop") %>%
  select(factor, Day, txt) %>%
  mutate(Day = paste0("Day ", Day)) %>%
  pivot_wider(names_from = Day, values_from = txt)
  
print(knitr::kable(dat_sn_stat_tbl, format = "markdown"))

```


# Plot - TV of each mouse {.tabset}
```{r plot_tv_each_mouse, fig.width = 6, fig.height = 8, results = "asis"}

dat_plot <- dat %>%
  mutate(factor = case_when(
    Dox.condition == "Dox" ~ paste0(genotype, " + dox"), 
    TRUE ~ paste0(genotype, " - dox")
  ))

factors <- unique(dat_plot$factor)

for (f in factors) {
  cat("\n##", f, "\n")
  dat_f <- dat_plot %>%
    filter(factor == f)
  gg_f <- dat_f %>%
    ggplot(aes(Day, TV, group = Animal.ID, label = round(TV, 0))) + 
    theme_classic() + 
    geom_point() + 
    geom_line() + 
    facet_wrap(~ Animal.ID, ncol = 4, scales = "free_y") + 
    geom_text_repel()
  print(gg_f)
  cat("\n")
  
}

```

